---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
if(!file.exists("activity.csv")) {
  unzip(zipfile = "activity.zip")
}
data <- read.csv("activity.csv")
data$date <- as.Date(data$date)
```

## What is mean total number of steps taken per day?
```{r, echo=FALSE}
steps_total <- aggregate(steps ~ date, data = data, sum, na.rm=TRUE)
hist(
  steps_total$steps,
  xlab = "Total number of steps per date",
  main = "Histogram of the total number of steps taken each day"
)
```
```{r, echo=FALSE}
options(scipen = 4)
mean_steps <- round(mean(steps_total$steps))
median_steps <- round(median(steps_total$steps))
```
Mean total number of steps taken per day: `r mean_steps`  
Median total number of steps taken per day: `r median_steps`  

## What is the average daily activity pattern?
```{r, echo=FALSE}
interval_average <- aggregate(steps ~ interval, data = data, mean, na.rm=TRUE)
plot(
  interval_average$interval, interval_average$steps, type="l",
  main = "Time series plot of the average number of steps taken",
  xlab = "Interval",
  ylab = "Average number of steps"
)
```
```{r, echo=FALSE}
max_interval <- interval_average$interval[which.max(interval_average$steps)]
```
The 5-minute interval that, on average, contains the maximum number of steps: `r max_interval`  

## Imputing missing values
```{r, echo=FALSE}
miss_count <- length(data$steps[is.na(data)])
```
The total number of missing values in the dataset is: `r miss_count`  
We create a new dataset, where missing values are replaced with the average number of steps for that interval across all days.  
```{r}
imputed_data <- data
imputed_data$steps[is.na(imputed_data$steps)] <- interval_average$steps[match(imputed_data$interval[is.na(imputed_data$steps)], interval_average$interval)]
```
```{r, echo=FALSE}
imputed_steps_total <- aggregate(steps ~ date, data = imputed_data, sum, na.rm=TRUE)
hist(
  imputed_steps_total$steps,
  xlab = "Total number of steps per date",
  main = "Histogram of the total number of steps taken each day (imputed data)"
)
```
```{r, echo=FALSE}
options(scipen = 4)
imputed_mean_steps <- round(mean(imputed_steps_total$steps))
imputed_median_steps <- round(median(imputed_steps_total$steps))
```
Mean total number of steps taken per day (imputed data): `r imputed_mean_steps`  
Median total number of steps taken per day (imputed data): `r imputed_median_steps`  
The results did not change much after imputing (only the median value increased slightly).
That might be due to the fact that we used average value of each interval for imputing.  

## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=FALSE}
imputed_data$weekday <- factor(ifelse(weekdays(imputed_data$date) %in% c('Saturday','Sunday'),'weekend','weekday'))
interval_average_weekday <- aggregate(steps ~ interval, data = subset(imputed_data, weekday=="weekday"), mean)
interval_average_weekend <- aggregate(steps ~ interval, data = subset(imputed_data, weekday=="weekend"), mean)
par(mfrow=c(2,1))
plot(
  interval_average_weekday$interval, interval_average_weekday$steps, type="l",
  main = "Weekdays",
  xlab = "Interval",
  ylab = "Average number of steps"
)
plot(
  interval_average_weekend$interval, interval_average_weekend$steps, type="l",
  main = "Weekend",
  xlab = "Interval",
  ylab = "Average number of steps"
)
```
